### 2014-02-28 paper/concrete.tex
concrete

The operational semantics we have defined in the previous section
satisfy non-interference by design.
We achieve this general statement that works for a large class of
languages by having different task execute completely isolated from
each other, such that every task has it's own state.
In some cases, this strong separation is desirable, or even necessary.
Languages like C provide direct access to the addresses without
mechanisms in the language to achieve a separation of the heap.
However, for other languages this isolation of tasks can be
undesirable, e.g., for performance reasons.
For instance, for the mini-ML language, our presentation so far
requires a separate heap per task, which is not very practical.
Instead, we would like to
more tightly couple the integration of the target and IFC
language by reusing existing infrastructure.  In the running example,
a concrete implementation might use a single global heap.
More precisely, instead of using a configuration of the form
\[|iconf iS (fullconf id1 il1 tS1 ie1, fullconf id2 il2 tS2 ie2 ldots)|\]
we would like a single global heap as in
|oneheapiconf iS tS (oneheapfullconf id1 il1 ie1, oneheapfullconf id2 il2 ie2, ldots)|.

If the operational rules are adapted na\"ively to this new setting,
then non-interference can be violated.  In particular, it is not
sound to share references to a heap cell between different tasks
in general.  Otherwise, information can be trivially leaked by copying
secret data into a reference in one task, and then
read again by a another task with a lower label.
Clearly, we cannot
modify the semantics in arbitrary ways if we are hoping to preserve
non-interference, but how can we know if our changes to the abstract
semantics preserve non-interference?
The intention of our single heap implementation was to be more efficient
by using one global heap, but conceptually maintain isolation between
task by not allowing sharing of references between tasks.
This intuition of having a different (potentially more efficient)
concrete semantics that behave like the abstract semantics
can be formalized by the following definition:

\begin{definition}[Isomorphism of information-flow control languages]
  A language \con{|(C, .->, erasef l)|} is \textit{isomorphic} to a
  language \abs{|(iC, .->, erasef l)|} if there exist total functions |f
  : tC -> iC| and |finv : iC -> tC| such that |f .. finv = idf tC| |finv
  .. f = idf iC|.  Furthermore, |f| and |finv| are functorial (e.g. if
  $\abs{x\ R\ y}$ then $f(\abs{x})\ \con{R}\ f(\abs{y})$) over both
  $l$-equivalences and |.->|.

@@

A na\"ive attempt at our single heap semantics
will not allow us to construct an isomorphism
with the abstract language |specLangML alpha|, and instead highlights
where things will go wrong.
To successfully construct an isomorphism, our concrete semantics need
to mirror the behavior of the abstract semantics, most importantly
in the case where tasks interact.  Namely, at a fork and when sending
or receiving messages, we take an expression from one task and
execute it in the context of another task.  In the abstract semantics
the fork rule is parameterized by the strategy |klone| which defines
what heap the new task should execute in.  \Red{In Section~\ref{sec:retrofit}
we have motivated two choices}, namely the identity (to effectively
clone the heap), or to start with an empty heap.
In the latter case, this effectively means that existing addresses
in the forked expression |ie| are no longer valid and the
task gets stuck (and then removed by \textsc{I-noStep}).
\todo{}{Is this actually true?  Currently the semantics are unclear.}
To get the isomorphic behavior in the concrete semantics with a
single heap, we would have to change the forked expression |ie| in
such a way that all addresses get rewritten to fresh addresses
that are guaranteed to not occur in the global heap.  Similarly,
sending a memory address should take a similar approach, as addresses
for sent references will not exist in the receiving task.

@@

Again we are modifying the IFC language semantics in a non-trivial way,
which begs the question whether non-interference is preserved.
Note that restricting the transitions in an information flow control
system in general is not non-interference preserving as the following
minimal example illustrates.
Consider a trivial state machine as shown
in Figure~\ref{fig:trivial-sm}, whose second projection is classified
secret.  With all three transitions, this language fulfills
termination sensitive non-interference.  Furthermore, if the
dashed transition
is removed, the language continues to satisfy TSNI.  However, if any
solid line is removed, the language fails TSNI.

@@

\begin{definition}[Restricted IFC language]
  \label{def:npri}
  For a family of predicates $\mathcal P$ (one for every reduction
  rule), we call
  |restrictedLang alpha targetLang| a restricted IFC language
  if its definition is equivalent to the abstract language
  |specLang alpha targetLang|, with the following exception:
  The reduction rules are restricted
  by adding a predicate $P$ from $\mathcal P$ to the premise of
  any rule other than \textsc{I-noStep}.  Furthermore, the predicate $P$
  can depend only on the \textit{erased} configuration
  |erase il ic|, where |il| is the label of the first task
  in the task list and |ic| the full configuration.
\end{definition}

@@

We are now ready to make our single heap IFC language precise and
ensure its non-interference using the techniques presented.
First, we can construct the restricted language
|restrictedLangNoRef alpha targetLangML|, where |noRefs| is
the family of always valid predicates, except for the ones for
\textsc{I-fork} and \textsc{I-send}, which we define as
\[ P = \text{|ie| does not contain any address |ta|} \]
Since $P$ only depends on |ie|, which is part of the current
task and thus never erased w.r.t.\ the label of the first task,
this language satisfies non-interference by Theorem~\ref{thm:restricted}.

The essential parts of the semantics for the concrete language
with a single heap are given in Figure~\Red{X}.  Most rules are
straight-forward translations of the rules in Figures~\ref{fig:ifc}
and~\ref{fig:embedding} but for a single heap.  For conciseness, we
only show the interesting ones.
Now, we can show an isomorphism between this language and
|restrictedLangNoRef alpha targetLangML|, which
(by Theorem~\ref{thm:iso-tsni} and~\ref{thm:iso-tini}) guarantees
non-interference for an appropriate scheduling policy |alpha|.

@@

\subsection{Old Stuff}

In this section, we will detail an approach that allows one to implement
a concrete system guided conceptually by the ideas in the previous
section but making different choice for ease of implementation or
performance, and give conditions for when such a system preserves
non-interference.
We first explain the idea for the single heap mini-ML language, and then
generalize to arbitrary languages.

\subsection{IFC-Mini-ML with a Single Heap}

To keep up a fictional separation between the portions
of the global heap that tasks can use, we forbid references to
cross task boundaries.  To achieve such a system, we can
adapt most of the operational rules easily to work with these
new configurations.  In the abstract system, the \textsc{I-fork}
rule created a fresh heap for the new task to execute in.  Since we
want to forbid references to cross task boundaries, we do not modify
the global heap and instead add an additional premise that the forked
expression |ie| must not contain any references.
Finally, we add the same condition to the \textsc{I-send} rule.
Tasks that try to fork or send an expression that contains a
reference will now get stuck, and will stop executing by the
\textsc{I-noStep} rule.
We call this family of languages |concreteLangMl alpha|, which
is still parametrized by the scheduling policy |alpha|.

\Red{Show the key rules here.}

In |concreteLangMl alpha| we
effectively restrict the abstract semantics in some cases (namely
to not allow the passing of references) as well as using a different
way to represent the heap.  How can we be sure that these
new, \textit{concrete semantics} for our system still satisfy
non-interference?

We will use two techniques to ensure non-interference.  First,
we require some form of isomorphism to the abstract language
|specLangML alpha| that ensures that the concrete system behaves
in a similar way.  Secondly, we pose conditions on what kind
of restrictions the concrete language is allowed to make
on the abstract language.

Intuitively, there are two scenarios in which non-interference
can be violated when restricting the transition relation:
\begin{itemize}
  \item Adding an additional premise to one of the rules that inspects
  information which is above the security label of the task currently
  executing.  This would allow another task to observe the termination
  of the other if both are at the same security level, and thus observe
  the secret information that was used in the additional
  premise of the rule.
  \item Adding additional restrictions to transition rules can
  also cause the whole configuration to get stuck, which can leak
  information through the termination channel (which is only
  of interest for TSNI, but not TINI, of course).
\end{itemize}

Next, we will make these two intuitions precise and give a general
definition for how non-interference can be preserved.  Then, we
show that the language |concreteLangMl alpha| satisfies these
conditions.

\Red{Should we give an example why removing transitions in general
  is not OK?  Edward has already typeset one that we could
  use.}

%This isn't trivial, as it is not sound to add arbitrary additional
%conditions on evaluation rules.  To see this, consider a contrived
%concrete system where we add a premise to a reduction rule that
%stops a task based on secret information.  Such a system will
%not satisfy non-interference, as it allows other tasks on the
%same level to observe the termination of that task and thus the
%secret information which was used to decide if the task terminates
%or not.

\subsection{Correctness of Concrete Semantics}

To formalize the conditions under which we can show the concrete
semantics preserve non-interference, we require some notation:

\begin{definition}[Isomorphism of information-flow control languages]
  A language \con{|(C, .->, erasef l)|} is \textit{isomorphic} to a
  language \abs{|(iC, .->, erasef l)|} if there exist total functions |f
  : tC -> iC| and |finv : iC -> tC| such that |f .. finv = idf tC| |finv
  .. f = idf iC|.  Furthermore, |f| and |finv| are functorial (e.g. if
  $\abs{x\ R\ y}$ then $f(\abs{x})\ \con{R}\ f(\abs{y})$) over both
  $l$-equivalences and |.->|.
\end{definition}

\Red{In some cases, this restriction can be weakened, where |finv| does
not have to be functorial over |.->|}

Now, we can propose the following two-step process to ensure that
non-interference is preserved:
\Red{We should find a better name for this, or not use a name, or something.}
\begin{definition}[Non-interference preserving, restricted isomorphism (NPRI)]
  \label{def:npri}
  We call an IFC language
  |concreteLang alpha targetLang| (the concrete language for a given
  target language |targetLang| and scheduling policy |alpha|)
  a \textit{non-intereference preserving,
  restricted isomorphism} with regard to the IFC language
  |specLang alpha targetLang|
  (the abstract language) if the following
  two conditions hold:
  \begin{enumerate}
    \item Any of the reduction rules of the abstract
    semantics |specLang alpha targetLang|
    presented in Section~\ref{sec:retrofit} can
    be restricted by adding a predicate $P$ to the premise of
    any rule other than \textsc{I-noStep}.  The predicate $P$
    can depend only on the \textit{erased} configuration
    |erase il ic|, where |il| is the label of the first task
    in the task list and |ic| the full configuration.
    \item The concrete language
    |concreteLang alpha targetLang|
    must be isomorphic to the abstract semantics |specLang alpha targetLang|
    with the additional predicates $P$ on the rules.
  \end{enumerate}
\end{definition}

\begin{theorem}[NPRI implies non-interference]
  \label{thm:npri}
  If a language |concreteLang alpha targetLang| is a NPRI, and |alpha|
  is a scheduling policy that makes |specLang alpha targetLang| satisfy
  TINI or TSNI, then also
  |concreteLang alpha targetLang| satisfies TINI, or TSNI, respectively.
\end{theorem}

Before we give the proof of this theorem in Section~\ref{sec:proofs},
let's consider our single-heap ML-like IFC language and
show that it is in fact NPRI, and thus non-interfering.

\begin{theorem}
  The language |concreteLangMl alpha| is a NPRI.
\end{theorem}
As a trivial corollary, it follows that |concreteLangMl roundrobinf| is
TSNI, and |concreteLangMl seqf| is TINI.

\begin{proof}
  It is easy to see that condition (1) of NPRI is satisfied, since
  we only add a side condition to the rules \textsc{I-fork} and
  \textsc{I-send}, and the condition only inspects the expression
  of the current task, which is by the definition of |erasef il| not
  erased.

  This leaves us to give an isomorphism between |concreteLangMl alpha|
  and |concreteLang alpha targetLangMl|, which we do as follows.
  \Red{Actually give an isomorphism.}
\end{proof}

\subsection{Proof of Theorem~\ref{thm:npri}}
\label{sec:proofs}

@@

\subsection{\Red{Previous Stuff (to be merged/removed as necessary)}}

However, modifying operational rules poses a distinct problem for
preserving non-interference.  Consider a trivial state machine as shown
in Figure~\ref{fig:trivial-sm}, whose second projection is classified
secret.  With all three transitions, this language fulfills
termination-sensitive noninterference.  Furthermore, if the dashed line
is removed, the language continues to satisfy TSNI.  However, if any
solid line is removed, however, the language fails TSNI.

Before we show isomorphism preserves termination sensitive
non-interference, we'd like to remark that one can safely remove transitions
and preserve termination \emph{insensitive} interference:

\begin{mathpar}
\inferrule[I-stepT]
{|
conf tS (te) -> conf tS' te'
|}
{|
coconf iS tS (cconf id il (iniE (IT te)), ldots)
.->
iS ; tS' ; sched step (cconf id il (iniE (IT te')), ldots)
|}

\and
% This rule is not complete, needs the FV condition
\inferrule[C-fork]
{ 
|iS' = iS [ id' mapsto nil ]|\\
|it1 = cconf id il (iniE id')|\\
|itnew = cconf id' il (TI ie)|\\
|fresh (id')|
}
{|
coconf iS tS (lconf id il (iniE (fork ie)), ldots)
.->
iS'; tS; sched fork (it1, ldots, itnew)
|}
\end{mathpar}

\caption{ML with a single heap}
\label{fig:comb}
\end{figure}

>>>>

The operational semantics we have defined in the previous section
satisfy non-interference by design.
We achieve this general statement that works for a large class of
languages by having different task execute completely isolated from
each other, such that every task has its own state.
In some cases, this strong separation is desirable, or even necessary.
Languages like C provide direct access to memory locations without
mechanisms in the language to achieve a separation of the heap.
On the other hand, for other languages this
strong isolation of tasks can be
undesirable, e.g., for performance reasons.
For instance, for the language |targetLangML|, our presentation so far
requires a separate heap per task, which is not very practical.
Instead, we would like to
more tightly couple the integration of the target and IFC
language by reusing existing infrastructure.  In the running example,
a concrete implementation might use a single global heap.
More precisely, instead of using a configuration of the form
\[|iconf iS (fullconf id1 il1 tS1 ie1, fullconf id2 il2 tS2 ie2 ldots)|\]
we would like a single global heap as in
\[|oneheapiconf iS tS (oneheapfullconf id1 il1 ie1, oneheapfullconf id2 il2 ie2, ldots)|\]

If the operational rules are adapted na\"ively to this new setting,
then non-interference can be violated.  In particular, it is not
sound to share references to a heap cell between different tasks.
Otherwise, information can be trivially leaked by copying
secret data into a reference in one task, and then
reading it again by a another task with a lower label.
Clearly, we cannot
modify the semantics in arbitrary ways if we are hoping to preserve
non-interference, but how can we know if our changes to the abstract
semantics preserve non-interference?
The intention of our single heap implementation was to be more efficient
by using one global heap, but conceptually maintain isolation between
task by not allowing sharing of references between tasks.
This intuition of having a different (potentially more efficient)
concrete semantics that behave like the abstract semantics
can be formalized by the following definition:

\Red{We really should not use red/blue here.  I suggest we just use primes instead of colors here.}
\begin{definition}[Isomorphism of information-flow control languages]
  A language \con{|(C, .->, erasef l)|} is \textit{isomorphic} to a
  language \abs{|(iC, .->, erasef l)|} if there exist total functions |f
  : tC -> iC| and |finv : iC -> tC| such that |f .. finv = idf tC| |finv
  .. f = idf iC|.  Furthermore, |f| and |finv| are functorial (e.g. if
  $\abs{x\ R\ y}$ then $f(\abs{x})\ \con{R}\ f(\abs{y})$) over both
  $l$-equivalences and |.->|.

@@

A na\"ive attempt at our single heap semantics
will not allow us to construct an isomorphism
with the abstract language |specLangML alpha|, and instead highlights
where things will go wrong.
To successfully construct an isomorphism, our concrete semantics need
to mirror the behavior of the abstract semantics, most importantly
in the case where tasks interact.  Namely, at a fork and when sending
or receiving messages, we take an expression from one task and
execute it in the context of another task.  In the abstract semantics
the fork rule is parameterized by the strategy |klone|, which defines
what heap the new task should execute in.  \Red{In Section~\ref{sec:retrofit}
we have motivated two choices}, namely the identity (to effectively
clone the heap), or to start with an empty heap.
In the latter case, this effectively means that existing addresses
in the forked expression |ie| are no longer valid and the
task gets stuck (and then removed by \textsc{I-noStep}).
\todo{}{Is this actually true?  Currently the semantics are unclear.}
To get the isomorphic behavior in the concrete semantics with a
single heap, we would have to change the forked expression |ie| in
such a way that all addresses get rewritten to fresh addresses
that are guaranteed to not occur in the global heap.  Similarly,
sending a memory address should take a similar approach, as addresses
for sent references will not exist in the receiving task.

@@

Again we are modifying the IFC language semantics in a non-trivial way,
which begs the question whether non-interference is preserved.
Note that restricting the transitions in an information flow control
system in general does not preserve non-interference as the following
minimal example illustrates.
Consider a trivial state machine as shown
in Figure~\ref{fig:trivial-sm}, whose second projection is classified
secret.  With all three transitions, this language fulfills
termination sensitive non-interference.  Furthermore, if the
dashed transition
is removed, the language continues to satisfy TSNI.  However, if any
solid line is removed, the language fails TSNI.

@@

\begin{definition}[Restricted IFC language]
  \label{def:npri}
  For a family of predicates $\mathcal P$ (one for every reduction
  rule), we call
  |restrictedLang alpha targetLang| a restricted IFC language
  if its definition is equivalent to the abstract language
  |specLang alpha targetLang|, with the following exception:
  The reduction rules are restricted
  by adding a predicate $P$ from $\mathcal P$ to the premise of
  all rule other than \textsc{I-noStep}.  Furthermore, the predicates $P$
  can depend only on the \textit{erased} configuration
  |erase il ic|, where |il| is the label of the first task
  in the task list and |ic| the full configuration.
\end{definition}

@@

We are now ready to make our single heap IFC language precise and
ensure its non-interference using the techniques presented.
First, we can construct the restricted language
|restrictedLangNoRef alpha targetLangML|, where |noRefs| is
the family of always valid predicates, except for the ones for
\textsc{I-fork} and \textsc{I-send}, which we define as
\[ P = \text{|ie| does not contain any address |ta|} \]
That is, we do not restrict any rules except for \textsc{I-fork}
and \textsc{I-send}.
Since $P$ only depends on |ie|, which is part of the current
task and thus never erased w.r.t.\ the label of the first task,
this language satisfies non-interference by Theorem~\ref{thm:restricted}.

  \begin{mathpar}
    \inferrule[C-fork]
    {
      \text{|ie| does not contain any |ta|}\\
      |iS' = iS [ id' mapsto nil ]|\\
      |it1 = oneheapfullconf id il (iniE id')|\\
      |itnew = oneheapfullconf id' il (TI ie)|\\
      |fresh (id')|
    }
    {|
      oneheapiconf iS tS (oneheapfullconf id1 il1 (iniEi (fork ie)), ldots)
      .->
      iS'; tS; sched F (it1, ldots, itnew)
    |}
    \and
    \inferrule[C-send]
    {
      \text{|ie| does not contain any |ta|}\\
      |il canFlowTo il'|\\
      |iS(id') = Q|\\
      |iS' = iS [ id' mapsto (il', id,  ie) , Q ]|
    }
    {|
      oneheapiconf iS tS (oneheapfullconf id il (iniEi (send id' il' ie)), ldots)
      ->
      iS; tS; sched step (oneheapfullconf id il unit, ldots)
    |}
  \end{mathpar}

  \caption{A selection of the reduction rules for |concreteLangMl alpha|.}
  \label{fig:concrete}
\end{figure}

The essential parts of the semantics for the concrete language
with a single heap,
which we call |concreteLangMl alpha|,
are given in Figure~\ref{fig:concrete}.  Most rules are
straight-forward translations of the rules in Figures~\ref{fig:ifc}
and~\ref{fig:embedding} but for a single heap.  For conciseness, we
only show the interesting ones.
Now, we can show an isomorphism between this language and
|restrictedLangNoRef alpha targetLangML|, which
(by Theorem~\ref{thm:iso-tsni} and~\ref{thm:iso-tini}) guarantees
non-interference for an appropriate scheduling policy |alpha|.

@@

\section{Proofs}
\label{sec:proofs}

\Red{Adapt some of this text.}

@@
